Working with daily climate model output data in R and the \pkg{futureheatwaves} package

### 2016-12-29 paper_drafts/RJournalRevisions/anderson-eason-barnes.tex
Add revisions to R Journal to git

---
title: >
    Working with daily climate model output data in R and the \pkg{futureheatwaves} package
author:
  - name: G. Brooke Anderson
    affiliation: Colorado State University
    address:
    - Department of Environmental & Radiological Health Sciences
    - 1681 Campus Delivery
    - Fort Collins, Colorado 80523
    email:  brooke.anderson@colostate.edu
  - name: Colin Eason
    affiliation: Colorado State University
    address:
    - Department of Computer Science
    - 1873 Campus Delivery
    - Fort Collins, Colorado 80523
    email:  aimesce@gmail.com
  - name: Elizabeth A. Barnes
    affiliation: Colorado State University
    address:
    - Department of Atmospheric Science
    - 1371 Campus Delivery
    - Fort Collins, CO 80523
    email:  eabarnes@atmos.colostate.edu
abstract: >
  Research on climate change impacts can require extensive processing, to incorporate output from ensembles of different climate models and different simulations of each model. This processing can be particularly extensive when identifying and characterizing extreme events like heat waves and frost day spells, as these must be processed from model output with daily time steps and may use identification threholds based on local distributions (e.g., 98th percentile temperature for that location). Here, we provide an overview of working with daily climate model output data in R. Further, we present the \pkg{futureheatwaves} package, which we developed to ease the process of identifying, characterizing, and exploring multi-day extreme events in climate model output. This package takes as input a directory of projection files from many climate models, identify all heat waves for multiple cities using a user-specified heat wave definition, and write an output directory of files with these heat waves and their characteristics (e.g., length, absolute and relative temperatures). The \pkg{futureheatwaves} package then allows the user to create custom functions and apply them across all heat wave files; this functionality can be used either to explore heat wave patterns (e.g., average length and intensity across different climate models) or to apply epidemiological estimates to project health impacts under different scenarios.
preamble: >
  % Supposedly allows you to add stuff to LaTeX preamble. Not for me.
output: rticles::rjournal_article
---

```{r echo = FALSE, message = FALSE, warning = FALSE}
library(futureheatwaves)
library(ggplot2)
```

# Introduction

Research on climate change impacts can require extensive processing. This is not only because output files for a single climate model can be large, but also because of the rising popularity of ensemble techniques \citep{IPCCch1}, in which, to better characterize uncertainty in projections, impacts are assessed for multiple climate models, multiple simulations of each climate model, and multiple climate experiments. Projections of regional climate change over the next century are subject to high uncertainty due to three distinct sources: (1) internal climate variability, i.e. climate noise, (2) climate model uncertainty, i.e. the same forcing can produce a different response in different models and (3) scenario uncertainty, i.e. uncertainty in future climate forcings (e.g. \citet{hawkins2009potential}). One approach for ensuring that these sources of uncertainty are characterized is to simulate the future climate many times with multiple models and for multiple future scenarios.

For example, the Coupled Model Intercomparison Project, phase 5 (CMIP5; \citet{taylor2012overview}) brought together dozens of major climate modeling groups around the world to simulate the same future radiative forcing scenarios, but with their own models. This created an ensemble of state-of-the-art climate model projections that allow researchers to study projections and their uncertainties. Most of these modeling groups additionally performed more than one simulation for each scenario and model (i.e. multiple ensemble members), perturbing the initial conditions by a very tiny amount to quantify uncertainties due to internal climate variability.

@@

The \pkg{futureheatwaves} package automates the process of identifying and characterizing multi-day extreme events across different ensemble members of one or more climate models. A variety of different heat wave definitions have been used to identify heat waves in a time series of temperature data \citep{smith2013heat}, and the choice of heat wave definitions can influence both health effect estimates \citep{chen2015influence, kent2014heat} and projected heat wave trends \citep{smith2013heat}. Further, other types of extreme events will be defined differently than heat waves (for example, frost day spells may be defined as one or more days with temperature at or below $0^oC$). This package therefore allows the user to create and use a custom the extreme event definition used to identify events in the climate model output. Finally, this package allows the user to explore the extreme events identified in the climate model output with a function that can take a user-defined R function and apply it across all extreme event files generated for the separate ensemble members.

# An overview of climate model output for R users

## CMIP5 climate model output data

@@

The CMIP5 climate model output data is distributed across data nodes at different climate modeling centers \citep{taylor2012overview}, but can be accessed centrally at the World Climate Research Programme CMIP5 data portal at https://pcmdi.llnl.gov/search/cmip5/. Users must register before downloading data, and some data are restricted to non-commercial use. There is a separate file for each combination of climate model, experiment, modeling realm (e.g., atmosphere, ocean), variable, time step, and ensemble member \citep{taylor2012overview, taylor2010cmip5}. For finer time scales, this output is further split across multiple files for specific year ranges (e.g., 5 years of output for each file) \citep{taylor2010cmip5}. Filenames for CMIP5 files can be parsed to generate information about the output variable, climate model, experiment, and ensemble member for the simulation \citep{taylor2010cmip5}. Files can be searched and downloaded through a point-and-click web interface. The data portal also allows you to generate a specific wget script, which you can use to download many files at once. A wget script can also be generated through Earth System Grid Federation's Search RESTful API. Tips on efficiently searching and downloading the data, including through use of wget scripts and the search API, are available as user tutorials through the website of the University of Colorado Boulder's Earth System CoG (e.g., https://www.earthsystemcog.org/projects/cog/doc/wget for a tutorial on downloading files using wget).

CMIP5 files are saved in Network Common Data Format (netCDF). NetCDF is a binary file format that allows storage of data representing a regular array. These climate model output files can be as large as several gigabytes [?]. For climate model output at a single depth (e.g., near-surface air temperature), the data is a 3-dimensional array, with dimensions representing time and two metrics of location (e.g., latitude and longitude) (Figure \ref{fig:netcdfexample}). Global climate models generate output at regularly-spaced time steps, typically at regularly-spaced grid points around the world. The latitude and longitude spacing of grid points vary by climate model, but are typically 1--2 degrees for atmospheric variables in CMIP5 models \citep{IPCCch9}. Each data point in the netCDF array gives the modeled value of the variable (e.g., surface temperature) for a single time point and location. For CMIP5 climate model output, the location units are in degrees east and degrees north for longitude and latitude, respectively. For daily output files, the time unit is in days since a specified origin date-time (e.g., days since 1850-01-01 00:00:00) \citep{taylor2010cmip5}. All CMIP5 output files are requred to include certain metadata  \citep{taylor2010cmip5}. This required metadata includes the experiment, forcing agents input to the model to create the simulation, time step, institution and institutional contact information, climate model, and modeling realm \citep{taylor2010cmip5}. The metadata also must include units for all of the coordinate variables (e.g., longitude, latitude, time).

@@

## Working with climate model output in R

@@

The netCDF format allows you to access metadata and variables describing the dimensions of the data without reading the full file into memory. The metadata describes the dimensions of the array and the units, etc., of each variable and is printed by the print method of the R object returned by the \code{nc\_open} function of \pkg{ncdf4}. For CMIP5 files, the dimension variables include the times and locations corresponding to each array in the data. Before reading in a variable from a netCDF file, a connection to the file must be opened, for example with the \code{nc\_open} function from \pkg{ncdf4}. Variables can then be read in using the \code{ncvar\_get} function from \pkg{ncdf4}, with the \code{varid} parameter set to "lat", "lon", or "time" (as a caveat, many climate models output to non-Gregorian calendars, in which case the time variable should be read in using a different function, as discussed later in this section). The climate output variable (e.g., near-surface air temperature) can similarly be read in using \code{ncvar\_get}. In this case, the \code{varid} parameter should be set using the appropriate CMIP5 variable name (e.g., "tas" for near-surface air temperature); these variable names can be found in the CMIP Requested Output tables \citep{taylor2010cmip5}. In practice, you can use the dimensional time and location data to identify the location of the variable data you need in the netCDF array and use indexing to read only that data into memory, without needing to read in the full file \citep{michna2013rnetcdf}. Once the user is done reading in data from the file, the connection can be closed (e.g., with the \code{nc\_close} function from \pkg{ncdf4}).

Since the late 1500s, Western dates have been set using the Gregorian calendar, which has 365.2425-day years. Some climate models, however, output to different calendars, including the Julian calendar (365.25-day years), a calendar where there are no leap years (365-day years), a calendar where every year is a leap year (366-day years), and a calendar of twelve 30-day months (360-day years) \citep{cfconventions}. With these non-Gregorian calendars, R's base functions for converting a vector to a Date class based on the number of days since an origin date (\code{as.Date}, \code{as.POSIXct}) do not return the desired values. The \pkg{PCICt} \citep{PCICt} and \pkg{ncdf4.helpers} \citep{ncdf4.helpers} packages provide further functionality with netCDF files that can be useful when working with climate model output data and provide particular help in working with different calendars. CMIP5 netCDF files include information on the calendar used, which the \code{nc.get.time.series} function in \pkg{ncdf4.helpers} use to convert the "time" variable in the file to an object of the PCICt class, which provides Date-like functionality for 360- and 365-day calendars \citep{PCICt}. As a note, while these functions will help with handling most CMIP5 files, the CMIP5 standards allows use of other calendars which may not be succesfully handled by these functions, so it is important to assess whether the time variable range in the PCICt object correctly matches the expected date ranges for a file as you process CMIP5 data in R.

@@

The following code gives an example of using a CMIP5 file and some of the R packages and functions discussed. Results include a map of near-surface air temperatures from a single climate model on a specific date (July 1, 2075) for China (Figure [x]) and a time series of daily near-surface air temperature simulations for the climate model grid point closest to Beijing, China (Figure [x]). To try this example, you will need to download the file ... from ... and save it in the "tmp" subdirectory of your home directory.

```{r}

```

@@

The \pkg{RCMIP5} package includes functions to read in CMIP5 data from netCDF files, scan a directory of CMIP5 files and determine models with continuous available data, create objects of a special "cmip5data" class to work with CMIP5 data within R, and parse the filenames for all files in a directory to extract information within the filename. For this package, most functions only work with monthly or less frequent data (e.g., the functions \code{checkTimePeriod} and \code{cmip5data}, as well as functions that work with \code{cmip5data} objects including \code{filterDimensions} and \code{getProjectionMatrix}) \citep{RCMIP5}. While the \code{loadCMIP5} function does successfully load the daily data as a \code{cmip5data} object, most of the methods for this object type do not do anything meaningful for daily data. The package's \code{getFileInfo} function, however, will work with CMIP5 files of any time step; this function identifies all CMIP5 files in a directory and creates a dataframe that parses the information contained in the file name. As a note, the \code{get.split.filename.cmip5} function in the \pkg{ncdf4.helper} package similarly can be used to parse information contained in CMIP5 file names \citep{ncdf4.helpers}.

The \pkg{wux} package \citep{wux} includes functions that allow the user to download CMIP5 monthly-aggregated output directly from within R with the \code{CMIP5fromESGF} function. However, this function does not allow downloading of climate model output with finer time steps, like daily data. This package uses the \code{models2wux} function to read in climate model output netCDF files and convert to a WUX dataframe that can be used by other functions in the package. While this function can input climate model output with daily time steps, if the element "what.timesteps" of the \code{modelinput} list input is set to "daily", the function aggregates this data to a monthly or less frequent (e.g., seasonal) aggregation when creating the WUX dataframe. Therefore, while this package provides very useful functionality for working with averaged output of daily climate model output data, it cannot easily be used to identify and characterize multi-day extreme events like heat waves.

More tips on working with CMIP5 files in R are provided in the "starting_from_netcdf" vignette of the \pkg{futureheatwaves} package.

# The \pkg{futureheatwaves} package

## How the package works

@@

The \pkg{futureheatwaves} package was facilitated to aid in identifying, characterizing, and exploring multi-day extreme events in daily climate model output data. Figure \ref{fig:overview} gives an overview of the two primary functions of the \pkg{futureheatwaves} package. First, the \code{gen\_hw\_set} function processes a directory of climate projection files that are stored locally on the user's computer (Figure \ref{fig:overview}, "Climate projections"), to generate a list of all extreme events in each projection, as well as over a dozen characteristics of each identified extreme event (Table \ref{tab:hwcharacteristics}). The extreme events are identified and characterized at one or more study locations (e.g., cities), with study locations specified by the user in an input file. The extreme events identified for each ensemble member are output as separate files in a directory specified by the user (Figure \ref{fig:overview}, ``Extreme events datasets"). Second, the \code{apply\_all\_models} function allows the user to apply custom functions across all the extreme event data frames generated by \code{gen\_hw\_set} to summaries of extreme events across all climate models and ensemble members (Figure \ref{fig:overview}, right).

@@

The user can extensively customize the definition used to identify extreme events. The default is a definition for heat waves (two or more days with temperatures at or above the city's $98^{th}$ percentile of year-round temperature), but users can write a custom R function with either a different heat wave definition or with a definition appropriate for a different type of extreme event (e.g., one or more days at or below $0^oC$ for frost day spells). 
Some of these heat wave characteristics are based on relative temperature, which is a measure of how the temperature during a heat wave compares to the typical distribution of temperature at that location \citep{IPCCch1}. These relative temperature metrics will vary depending on whether you calculate them based on a location's present-day temperature distribution or on the location's temperature distribution in the future, since typical temperature distributions in many locations are expected to change with climate change. The package therefore allows the user to specify date ranges of the temperature distributions to be used in calculating these relative temperature metrics in each location.

Second, once the user creates these data frames of location-specific heat waves, the `apply_all_models` function allows the user to apply a custom function across all the heat wave data frames that were generated to create a summary data frame (Figure \ref{fig:overview}, right). This functionality can be used to generate summary statistics (e.g., determine average heat wave length or total heat wave days), or can be used to apply more complex functions (e.g., apply epidemiologic effect estimates across the heat waves to generate health impact estimates).

## Example data

We have included data files in the package to serve as example files, so users can try this package before applying it to their own directory of climate projection files. The example files are all included as comma-separated (.csv) files, rather than as saved R objects, because the `gen_hw_set` function requires a directory of comma-separated files as input and csv is a general file format familiar to most users. This example data therefore also allow users to see how to create the appropriate directory set-up for this package.

@@

Once the \pkg{futureheatwaves} package is installed and loaded, the user can find the local location of these files using R's `system.file` function. In the later sections of this article, we show how to use the package functions with these example files as inputs.

When setting up to run climate model output data beyond this example data, any downloaded CMIP5 files will require some pre-processing, in terms of getting the getting the data into a specific format and storing these files in a specific directory structure. A file of study locations must also be saved locally in a specific format. Extensive details on the required format of files and required input directory structure is available in the \pkg{futureheatwaves} vignette "futureheatwaves", while tips on processing CMIP5 files into the required structure are given in the package's "starting_from_netcdf" vignette.

## Basic example of processing projection files

Once these files and directories are set up, \pkg{futureheatwaves} can process them to identify and characterize heat waves in each ensemble member's projection for each  location using the `gen_hw_set` function. For example, to process the example climate projections directory included with the package, the user can run:

```{r eval = FALSE}
projection_dir_location <- system.file("extdata/cmip5",
                                       package = "futureheatwaves")
city_file_location <- system.file("extdata/cities.csv",
                                  package = "futureheatwaves")

gen_hw_set(out = "example_results",
           dataFolder = projection_dir_location ,
           dataDirectories = list("historical" = c(1990, 1999),
                                        "rcp85" = c(2060, 2079)),
           citycsv = city_file_location,
           coordinateFilenames = "latitude_longitude_NorthAmerica_12mo.csv",
           tasFilenames = "tas_NorthAmerica_12mo.csv",
           timeFilenames = "time_NorthAmerica_12mo.csv")
```

This code first identifies and saves as objects the path names on the user's computer of the example climate projections directory (`projection_dir_location`) and the file of study locations (`city_file_location`). The `gen_hw_set` function processes this example input and creates a new directory, `example_results`, with files of identified and characterized heat waves, in the user's current working directory. In this example code, this processing is done using  default values for the heat wave definition, years for which to generate the heat wave data sets, etc. Ways to customize these choices are fully explained later in the text.

In this call, the user must specify the directory where the results should be written (`out`), the location of the directory of climate projections (`dataFolder`), the names of the two main subdirectories of that climate projection directory, as well as their year boundaries (`dataDirectories`; see, for example, Figure \ref{fig:directories}), the location of the city location file (`citycsv`), and the names used for the grid coordinate, climate projection, and projection date files (`coordinateFilenames`, `tasFilenames`, and `timeFilenames`). When `gen_hw_set` is run, the user is advised that the function will write files to his computer and must agree to proceed:

```
 Warning: This function will write new files to your computer in the 
 ~/tmp/example_results/ directory of your computer. If that directory already exists,
running this function will write over it. 
 Do you want to continue? (y / n): 
```

\noindent The function provides status reports while processing. This helps the user see that the function call is progressing, as `gen_hw_set` can take a while to run if it is processing many locations and / or many climate ensemble members. Once the function has completed running, results will be written locally to the directory specified by the `out` argument of `gen_hw_set`. The structure of this output directory is shown in Figure \ref{fig:directories} (right). This directory will include files with some basic information about the climate models and the closest grid points of each climate model to each location. The output directory will also include a directory with files of identified and classified heat waves for each ensemble member, including all characteristics in Table \ref{tab:hwcharacteristics}.

## Customizing the extreme event definition

@@

> A \dfn{heat wave} is two or more days at or above a city-specific threshold temperature, with the threshold determined as the $98^{th}$ percentile of year-round temperature in the city during some reference period (by default, 1990--1999).

Many different definitions of heat waves exist (e.g., \citet{smith2013heat, kent2014heat}), so researchers will often want to use alternative ways to define heat waves. Researchers might want to use a specific definition, for example, because it matches the definition used by local health officials to declare heat wave warnings or, in the case of health impact assessments, to match with a definition used in an epidemiological study. Three components of the heat wave definition can be easily customized in the `gen_hw_set` function call, without creating a new R function to use to identify heat waves. The customization of the heat wave definition is even more extensive as one has the option of writing a custom R function.

>>>>

% !TeX root = RJwrapper.tex
\title{Working with daily climate model output data in R and the
\pkg{futureheatwaves} package}
\author{by G. Brooke Anderson, Colin Eason, Elizabeth A. Barnes}

\maketitle

\abstract{%
Research on climate change impacts can require extensive processing, to
incorporate output from ensembles of different climate models and
different simulations of each model. This processing can be particularly
extensive when identifying and characterizing extreme events like heat
waves and frost day spells, as these must be processed from model output
with daily time steps and may use identification threholds based on
local distributions (e.g., 98th percentile temperature for that
location). Here, we provide an overview of working with daily climate
model output data in R. Further, we present the \pkg{futureheatwaves}
package, which we developed to ease the process of identifying,
characterizing, and exploring multi-day extreme events in climate model
output. This package takes as input a directory of projection files from
many climate models, identify all heat waves for multiple cities using a
user-specified heat wave definition, and write an output directory of
files with these heat waves and their characteristics (e.g., length,
absolute and relative temperatures). The \pkg{futureheatwaves} package
then allows the user to create custom functions and apply them across
all heat wave files; this functionality can be used either to explore
heat wave patterns (e.g., average length and intensity across different
climate models) or to apply epidemiological estimates to project health
impacts under different scenarios.
}

\section{Introduction}\label{introduction}

Research on climate change impacts can require extensive processing.
This is not only because output files for a single climate model can be
large, but also because of the rising popularity of ensemble techniques
\citep{IPCCch1}, in which, to better characterize uncertainty in
projections, impacts are assessed for multiple climate models, multiple
simulations of each climate model, and multiple climate experiments.
Projections of regional climate change over the next century are subject
to high uncertainty due to three distinct sources: (1) internal climate
variability, i.e.~climate noise, (2) climate model uncertainty, i.e.~the
same forcing can produce a different response in different models and
(3) scenario uncertainty, i.e.~uncertainty in future climate forcings
(e.g. \citet{hawkins2009potential}). One approach for ensuring that
these sources of uncertainty are characterized is to simulate the future
climate many times with multiple models and for multiple future
scenarios.

For example, the Coupled Model Intercomparison Project, phase 5 (CMIP5;
\citet{taylor2012overview}) brought together dozens of major climate
modeling groups around the world to simulate the same future radiative
forcing scenarios, but with their own models. This created an ensemble
of state-of-the-art climate model projections that allow researchers to
study projections and their uncertainties. Most of these modeling groups
additionally performed more than one simulation for each scenario and
model (i.e.~multiple ensemble members), perturbing the initial
conditions by a very tiny amount to quantify uncertainties due to
internal climate variability.

@@

The \pkg{futureheatwaves} package automates the process of identifying
and characterizing multi-day extreme events across different ensemble
members of one or more climate models. A variety of different heat wave
definitions have been used to identify heat waves in a time series of
temperature data \citep{smith2013heat}, and the choice of heat wave
definitions can influence both health effect estimates
\citep{chen2015influence, kent2014heat} and projected heat wave trends
\citep{smith2013heat}. Further, other types of extreme events will be
defined differently than heat waves (for example, frost day spells may
be defined as one or more days with temperature at or below \(0^oC\)).
This package therefore allows the user to create and use a custom the
extreme event definition used to identify events in the climate model
output. Finally, this package allows the user to explore the extreme
events identified in the climate model output with a function that can
take a user-defined R function and apply it across all extreme event
files generated for the separate ensemble members.

\section{An overview of climate model output for R
users}\label{an-overview-of-climate-model-output-for-r-users}

\subsection{CMIP5 climate model output
data}\label{cmip5-climate-model-output-data}

@@

The CMIP5 climate model output data is distributed across data nodes at
different climate modeling centers \citep{taylor2012overview}, but can
be accessed centrally at the World Climate Research Programme CMIP5 data
portal at \url{https://pcmdi.llnl.gov/search/cmip5/}. Users must
register before downloading data, and some data are restricted to
non-commercial use. There is a separate file for each combination of
climate model, experiment, modeling realm (e.g., atmosphere, ocean),
variable, time step, and ensemble member
\citep{taylor2012overview, taylor2010cmip5}. For finer time scales, this
output is further split across multiple files for specific year ranges
(e.g., 5 years of output for each file) \citep{taylor2010cmip5}.
Filenames for CMIP5 files can be parsed to generate information about
the output variable, climate model, experiment, and ensemble member for
the simulation \citep{taylor2010cmip5}. Files can be searched and
downloaded through a point-and-click web interface. The data portal also
allows you to generate a specific wget script, which you can use to
download many files at once. A wget script can also be generated through
Earth System Grid Federation's Search RESTful API. Tips on efficiently
searching and downloading the data, including through use of wget
scripts and the search API, are available as user tutorials through the
website of the University of Colorado Boulder's Earth System CoG (e.g.,
\url{https://www.earthsystemcog.org/projects/cog/doc/wget} for a
tutorial on downloading files using wget).

CMIP5 files are saved in Network Common Data Format (netCDF). NetCDF is
a binary file format that allows storage of data representing a regular
array. These climate model output files can be as large as several
gigabytes {[}?{]}. For climate model output at a single depth (e.g.,
near-surface air temperature), the data is a 3-dimensional array, with
dimensions representing time and two metrics of location (e.g., latitude
and longitude) (Figure \ref{fig:netcdfexample}). Global climate models
generate output at regularly-spaced time steps, typically at
regularly-spaced grid points around the world. The latitude and
longitude spacing of grid points vary by climate model, but are
typically 1--2 degrees for atmospheric variables in CMIP5 models
\citep{IPCCch9}. Each data point in the netCDF array gives the modeled
value of the variable (e.g., surface temperature) for a single time
point and location. For CMIP5 climate model output, the location units
are in degrees east and degrees north for longitude and latitude,
respectively. For daily output files, the time unit is in days since a
specified origin date-time (e.g., days since 1850-01-01 00:00:00)
\citep{taylor2010cmip5}. All CMIP5 output files are requred to include
certain metadata \citep{taylor2010cmip5}. This required metadata
includes the experiment, forcing agents input to the model to create the
simulation, time step, institution and institutional contact
information, climate model, and modeling realm \citep{taylor2010cmip5}.
The metadata also must include units for all of the coordinate variables
(e.g., longitude, latitude, time).

@@

\subsection{Working with climate model output in
R}\label{working-with-climate-model-output-in-r}

@@

The netCDF format allows you to access metadata and variables describing
the dimensions of the data without reading the full file into memory.
The metadata describes the dimensions of the array and the units, etc.,
of each variable and is printed by the print method of the R object
returned by the \code{nc\_open} function of \pkg{ncdf4}. For CMIP5
files, the dimension variables include the times and locations
corresponding to each array in the data. Before reading in a variable
from a netCDF file, a connection to the file must be opened, for example
with the \code{nc\_open} function from \pkg{ncdf4}. Variables can then
be read in using the \code{ncvar\_get} function from \pkg{ncdf4}, with
the \code{varid} parameter set to ``lat'', ``lon'', or ``time'' (as a
caveat, many climate models output to non-Gregorian calendars, in which
case the time variable should be read in using a different function, as
discussed later in this section). The climate output variable (e.g.,
near-surface air temperature) can similarly be read in using
\code{ncvar\_get}. In this case, the \code{varid} parameter should be
set using the appropriate CMIP5 variable name (e.g., ``tas'' for
near-surface air temperature); these variable names can be found in the
CMIP Requested Output tables \citep{taylor2010cmip5}. In practice, you
can use the dimensional time and location data to identify the location
of the variable data you need in the netCDF array and use indexing to
read only that data into memory, without needing to read in the full
file \citep{michna2013rnetcdf}. Once the user is done reading in data
from the file, the connection can be closed (e.g., with the
\code{nc\_close} function from \pkg{ncdf4}).

Since the late 1500s, Western dates have been set using the Gregorian
calendar, which has 365.2425-day years. Some climate models, however,
output to different calendars, including the Julian calendar (365.25-day
years), a calendar where there are no leap years (365-day years), a
calendar where every year is a leap year (366-day years), and a calendar
of twelve 30-day months (360-day years) \citep{cfconventions}. With
these non-Gregorian calendars, R's base functions for converting a
vector to a Date class based on the number of days since an origin date
(\code{as.Date}, \code{as.POSIXct}) do not return the desired values.
The \pkg{PCICt} \citep{PCICt} and \pkg{ncdf4.helpers}
\citep{ncdf4.helpers} packages provide further functionality with netCDF
files that can be useful when working with climate model output data and
provide particular help in working with different calendars. CMIP5
netCDF files include information on the calendar used, which the
\code{nc.get.time.series} function in \pkg{ncdf4.helpers} use to convert
the ``time'' variable in the file to an object of the PCICt class, which
provides Date-like functionality for 360- and 365-day calendars
\citep{PCICt}. As a note, while these functions will help with handling
most CMIP5 files, the CMIP5 standards allows use of other calendars
which may not be succesfully handled by these functions, so it is
important to assess whether the time variable range in the PCICt object
correctly matches the expected date ranges for a file as you process
CMIP5 data in R.

@@

The following code gives an example of using a CMIP5 file and some of
the R packages and functions discussed. Results include a map of
near-surface air temperatures from a single climate model on a specific
date (July 1, 2075) for China (Figure {[}x{]}) and a time series of
daily near-surface air temperature simulations for the climate model
grid point closest to Beijing, China (Figure {[}x{]}). To try this
example, you will need to download the file \ldots{} from \ldots{} and
save it in the ``tmp'' subdirectory of your home directory.

@@

The \pkg{RCMIP5} package includes functions to read in CMIP5 data from
netCDF files, scan a directory of CMIP5 files and determine models with
continuous available data, create objects of a special ``cmip5data''
class to work with CMIP5 data within R, and parse the filenames for all
files in a directory to extract information within the filename. For
this package, most functions only work with monthly or less frequent
data (e.g., the functions \code{checkTimePeriod} and \code{cmip5data},
as well as functions that work with \code{cmip5data} objects including
\code{filterDimensions} and \code{getProjectionMatrix}) \citep{RCMIP5}.
While the \code{loadCMIP5} function does successfully load the daily
data as a \code{cmip5data} object, most of the methods for this object
type do not do anything meaningful for daily data. The package's
\code{getFileInfo} function, however, will work with CMIP5 files of any
time step; this function identifies all CMIP5 files in a directory and
creates a dataframe that parses the information contained in the file
name. As a note, the \code{get.split.filename.cmip5} function in the
\pkg{ncdf4.helper} package similarly can be used to parse information
contained in CMIP5 file names \citep{ncdf4.helpers}.

The \pkg{wux} package \citep{wux} includes functions that allow the user
to download CMIP5 monthly-aggregated output directly from within R with
the \code{CMIP5fromESGF} function. However, this function does not allow
downloading of climate model output with finer time steps, like daily
data. This package uses the \code{models2wux} function to read in
climate model output netCDF files and convert to a WUX dataframe that
can be used by other functions in the package. While this function can
input climate model output with daily time steps, if the element
``what.timesteps'' of the \code{modelinput} list input is set to
``daily'', the function aggregates this data to a monthly or less
frequent (e.g., seasonal) aggregation when creating the WUX dataframe.
Therefore, while this package provides very useful functionality for
working with averaged output of daily climate model output data, it
cannot easily be used to identify and characterize multi-day extreme
events like heat waves.

More tips on working with CMIP5 files in R are provided in the
``starting\_from\_netcdf'' vignette of the \pkg{futureheatwaves}
package.

\section{\texorpdfstring{The \pkg{futureheatwaves}
package}{The  package}}\label{the-package}

\subsection{How the package works}\label{how-the-package-works}

@@

The \pkg{futureheatwaves} package was facilitated to aid in identifying,
characterizing, and exploring multi-day extreme events in daily climate
model output data. Figure \ref{fig:overview} gives an overview of the
two primary functions of the \pkg{futureheatwaves} package. First, the
\code{gen\_hw\_set} function processes a directory of climate projection
files that are stored locally on the user's computer (Figure
\ref{fig:overview}, ``Climate projections''), to generate a list of all
extreme events in each projection, as well as over a dozen
characteristics of each identified extreme event (Table
\ref{tab:hwcharacteristics}). The extreme events are identified and
characterized at one or more study locations (e.g., cities), with study
locations specified by the user in an input file. The extreme events
identified for each ensemble member are output as separate files in a
directory specified by the user (Figure \ref{fig:overview}, ``Extreme
events datasets``). Second, the \code{apply\_all\_models} function
allows the user to apply custom functions across all the extreme event
data frames generated by \code{gen\_hw\_set} to summaries of extreme
events across all climate models and ensemble members (Figure
\ref{fig:overview}, right).

@@

The user can extensively customize the definition used to identify
extreme events. The default is a definition for heat waves (two or more
days with temperatures at or above the city's \(98^{th}\) percentile of
year-round temperature), but users can write a custom R function with
either a different heat wave definition or with a definition appropriate
for a different type of extreme event (e.g., one or more days at or
below \(0^oC\) for frost day spells). Some of these heat wave
characteristics are based on relative temperature, which is a measure of
how the temperature during a heat wave compares to the typical
distribution of temperature at that location \citep{IPCCch1}. These
relative temperature metrics will vary depending on whether you
calculate them based on a location's present-day temperature
distribution or on the location's temperature distribution in the
future, since typical temperature distributions in many locations are
expected to change with climate change. The package therefore allows the
user to specify date ranges of the temperature distributions to be used
in calculating these relative temperature metrics in each location.

Second, once the user creates these data frames of location-specific
heat waves, the \texttt{apply\_all\_models} function allows the user to
apply a custom function across all the heat wave data frames that were
generated to create a summary data frame (Figure \ref{fig:overview},
right). This functionality can be used to generate summary statistics
(e.g., determine average heat wave length or total heat wave days), or
can be used to apply more complex functions (e.g., apply epidemiologic
effect estimates across the heat waves to generate health impact
estimates).

\subsection{Example data}\label{example-data}

We have included data files in the package to serve as example files, so
users can try this package before applying it to their own directory of
climate projection files. The example files are all included as
comma-separated (.csv) files, rather than as saved R objects, because
the \texttt{gen\_hw\_set} function requires a directory of
comma-separated files as input and csv is a general file format familiar
to most users. This example data therefore also allow users to see how
to create the appropriate directory set-up for this package.

@@

Once the \pkg{futureheatwaves} package is installed and loaded, the user
can find the local location of these files using R's
\texttt{system.file} function. In the later sections of this article, we
show how to use the package functions with these example files as
inputs.

When setting up to run climate model output data beyond this example
data, any downloaded CMIP5 files will require some pre-processing, in
terms of getting the getting the data into a specific format and storing
these files in a specific directory structure. A file of study locations
must also be saved locally in a specific format. Extensive details on
the required format of files and required input directory structure is
available in the \pkg{futureheatwaves} vignette ``futureheatwaves'',
while tips on processing CMIP5 files into the required structure are
given in the package's ``starting\_from\_netcdf'' vignette.

\subsection{Basic example of processing projection
files}\label{basic-example-of-processing-projection-files}

Once these files and directories are set up, \pkg{futureheatwaves} can
process them to identify and characterize heat waves in each ensemble
member's projection for each location using the \texttt{gen\_hw\_set}
function. For example, to process the example climate projections
directory included with the package, the user can run:

\begin{Schunk}
\begin{Sinput}
projection_dir_location <- system.file("extdata/cmip5",
                                       package = "futureheatwaves")
city_file_location <- system.file("extdata/cities.csv",
                                  package = "futureheatwaves")

gen_hw_set(out = "example_results",
           dataFolder = projection_dir_location ,
           dataDirectories = list("historical" = c(1990, 1999),
                                        "rcp85" = c(2060, 2079)),
           citycsv = city_file_location,
           coordinateFilenames = "latitude_longitude_NorthAmerica_12mo.csv",
           tasFilenames = "tas_NorthAmerica_12mo.csv",
           timeFilenames = "time_NorthAmerica_12mo.csv")
\end{Sinput}
\end{Schunk}

This code first identifies and saves as objects the path names on the
user's computer of the example climate projections directory
(\texttt{projection\_dir\_location}) and the file of study locations
(\texttt{city\_file\_location}). The \texttt{gen\_hw\_set} function
processes this example input and creates a new directory,
\texttt{example\_results}, with files of identified and characterized
heat waves, in the user's current working directory. In this example
code, this processing is done using default values for the heat wave
definition, years for which to generate the heat wave data sets, etc.
Ways to customize these choices are fully explained later in the text.

In this call, the user must specify the directory where the results
should be written (\texttt{out}), the location of the directory of
climate projections (\texttt{dataFolder}), the names of the two main
subdirectories of that climate projection directory, as well as their
year boundaries (\texttt{dataDirectories}; see, for example, Figure
\ref{fig:directories}), the location of the city location file
(\texttt{citycsv}), and the names used for the grid coordinate, climate
projection, and projection date files (\texttt{coordinateFilenames},
\texttt{tasFilenames}, and \texttt{timeFilenames}). When
\texttt{gen\_hw\_set} is run, the user is advised that the function will
write files to his computer and must agree to proceed:

\begin{verbatim}
 Warning: This function will write new files to your computer in the 
 ~/tmp/example_results/ directory of your computer. If that directory already exists,
running this function will write over it. 
 Do you want to continue? (y / n): 
\end{verbatim}

\noindent The function provides status reports while processing. This
helps the user see that the function call is progressing, as
\texttt{gen\_hw\_set} can take a while to run if it is processing many
locations and / or many climate ensemble members. Once the function has
completed running, results will be written locally to the directory
specified by the \texttt{out} argument of \texttt{gen\_hw\_set}. The
structure of this output directory is shown in Figure
\ref{fig:directories} (right). This directory will include files with
some basic information about the climate models and the closest grid
points of each climate model to each location. The output directory will
also include a directory with files of identified and classified heat
waves for each ensemble member, including all characteristics in Table
\ref{tab:hwcharacteristics}.

\subsection{Customizing the extreme event
definition}\label{customizing-the-extreme-event-definition}

@@

\begin{quote}
A \dfn{heat wave} is two or more days at or above a city-specific
threshold temperature, with the threshold determined as the \(98^{th}\)
percentile of year-round temperature in the city during some reference
period (by default, 1990--1999).
\end{quote}
